A self-consistency check for unitary propagation of Hawking quanta

### 2016-11-28 Draft_BH_shells.tex

\begin{abstract}
The black hole information paradox presumes that quantum field theory in curved spacetime can provide unitary propagations from near-horizon modes to asymptotic Hawking quanta. 
Instead of invoking conjectural quantum-gravity effects to modify such an assumption, we propose a self-consistency check.
We establish an analogy to Feynman's analysis of a double-slit experiment. 
Feynman showed that unitary propagation of the interfering particles, namely ignoring the entanglement with the double-slit, becomes an arbitrarily reliable assumption when the screen upon which the interference pattern is projected is infinitely far away.
We argue for an analogous self-consistency check for quantum field theory in curved spacetime.
We apply it to the propagation of Hawking quanta and test whether ignoring the entanglement with the geometry also becomes arbitrarily reliable in the limit of a large black hole. 
We present curious results to suggest a negative answer, and we discuss how this loss of na\"ive unitarity in QFT might be related to a solution of the paradox based on the soft-hair-memory effect.
\end{abstract}

@@

The black hole information paradox \cite{Haw76a} was sharpened to show a self-inconsistency among the following three widely-believed statements \cite{AMPS}\footnote{Different physicists may believe in some of those more strongly than the others, but very few would have argued against any single statement without the explicit conflict in the information paradox.}:
\begin{itemize}
\item {\bf Unitary evaporation:} A black hole will totally evaporate; the formation and evaporation of a black hole is described by an asymptotic observer as a unitary $S$-matrix, whose size is the exponential of the black hole's Bekenstein entropy.
\item {\bf General relativity:} The collapse-Schwarzschild geometry given by general relativity is a valid description of the spacetime everywhere except for near the singularity.
Therefore, there is no drama while crossing the horizon.
\item {\bf Quantum field theory in curved spacetime:} Away from the singularity, we can apply local quantum field theory (QFT) which describes microscopic steps of evaporation. 
For large black holes, it manifests itself as unitary propagation of every near-horizon mode into every asymptotic Hawking quanta. The back-reaction of such effect is no more than a Schwarzschild metric with a very slowly changing mass.
\end{itemize}
Basically, Statement Three demands that a near horizon mode is related to an asymptotic Hawking quanta by a unitary transformation, and therefore carries the same qubit of information. 
However, Statement One and Two demand this unique qubit to be maximally entangled with two distinct objects, violating monogamy \cite{Bou13}. 
This conflict has inspired many different proposals to modify one of the above three statements. 
For example, information loss or remnant \cite{Bek94} modifies Statement One; 
firewall \cite{BraPir09,AMPS}, ER=EPR \cite{MalSus13}, pre-Hawking radiation \cite{KawMat13,Ho16}, or some version of the fuzzball \cite{Mat15} modifies Statement Two;
various proposals of nonlocal effects near the horizon \cite{Gid12,DodSil15,OsuPag16}, causal patch complementarity \cite{HuiYan13,IlgYan13,LowTho14}, or computational complementarity \cite{HH} modify Statement Three.\footnote{Various versions of complementarity basically claim that QFT breaks down if one applies it to compare quantities which are in-principle not measurable (or difficult to measure) by the same observer.}

In this paper, we will try to provide a stronger motivation to modify Statement Three. 
In fact, we will argue that it is already problematic on its own. 
In other words, {\bf the application of local QFT to unitary propagation of Hawking quanta is not self-consistent. }

Before presenting our argument, we will first explain its relation to some existing ideas.
The recent paper by Osuga and Page \cite{OsuPag16} provides an abstract framework of how modifying Statement Three resolves the paradox.
Basically, the near horizon mode is thermal as demanded by Statement Two.
However, it propagates out in a non-unitary process, gaining the appropriate information to restore the unitary $S$-matrix demanded by Statement One. 
For this abstract model to work in practice, this non-unitary propagation should be understood as an interaction with a hidden system.
Here hidden simply means something that a na\"ive application of local QFT is ignorant to.
The obvious candidate of such hidden system is the geometry.

One proposal along this line of thoughts was discussed in \cite{NomVar12}. 
The idea was that the black hole acquires a large uncertainty during the evaporation process. 
If we treat the superposition of different classical geometries as a quantum system, it may qualify as the hidden system that is responsible for the non-unitary propagation of Hawking quanta. 
Unfortunately, the early version of this proposal can be defeated by counting entropy.
In order for such non-unitary process to restore the asymptotic $S$-matrix, the hidden system must store the required information, which is comparable to the amount of microscopic information carried by the black hole.
However, the black-hole-no-hair theorem states that two black holes with identical macroscopic parameters (mass, momentum, angular momentum, charge) are isometric to each other. 
This limits the number of different classical geometries to be parametrized by macroscopic parameters only.
As a result, the Hilbert space of superposed classical geometries has far fewer dimensions than what is necessary to carry the required amount of information.\footnote{We thank Raphael Bousso for private communication.}
Thus, even if a propagating Hawking quantum does interact with this hidden system, it cannot reproduce a unitary $S$-matrix; therefore such modification of Statement Three alone is insufficient to resolve the paradox.

The recent development on black-hole-soft-hairs \cite{HawPer16} hinted a way to revive this idea.
Essentially, it was realized that two isometric (regions of) geometries should not always be considered as the same state.
If one maps the geometries with congruences of geodesics, two isometric regions with different maps are actually different states.
That is because changes in the mapping geodesics are measurable memory effects, and a consistent physical theory must be able to distinguish them.
This realization may dramatically increase the number of different states encoded in classical geometries.
A tentative counting of state on a black hole horizon yields exactly the required number such that a modification to Statement Three can indeed resolve the paradox \cite{HawPer16}.

In this paper, we will take another step that is complementary to the above progress. It is exciting to know that modifying Statement Three can, in-principle, resolve the paradox, yet it remains a great mystery why local QFT should break down in low curvature regions. 
If we just demand that it breaks down in order to resolve the information paradox, then there is no reason why this solution is preferred over modifying Statement One or Two. 
We would like to argue that there is a self-consistency check that we should apply to QFT in curved spacetime, for a reason that is totally independent from the information paradox.
Applying such self-consistency check, we can show that a unitary propagation from a near horizon mode to an asymptotic Hawking quantum is already self-inconsistent. 
Thus, it is only natural to modify Statement Three.

@@

Fortunately, Feynman, in his famous lectures, presented an interesting trick to circumvent this obstacle. 
This trick, to a certain extent, enables us to perform a {\bf self-consistency check} of the classical-background assumption {\bf without} knowing the quantum nature of the background.
In Sec.\ref{sec-DoubleSlit}, we will review Feynman's analysis of the double-slit experiment. 
In his case, the double-slit is the classical background, the particle passing through it is the quantum subsystem, and subsystem unitarity is checked by whether an interference pattern can be observed on the final projection screen. 
He showed that we can ignore the quantum details of the double-slit and summarize that as an uncertainty in its position, which is a classical quantity. 
When such uncertainty is fixed, a classical calculation can show that the interference pattern is always visible as long as we put the final projection screen to be infinitely far away. 
It is in this sense that a unitary evolution of the subsystem is an arbitrarily good approximation.

The fundamental principle behind Feynman's trick is that ``classical wave coherence'', in his case the observability of the interference pattern, should be taken as the prerequisite of ``quantum unitarity''. 
In Sec.\ref{sec-QFT}, we argue that this principle enables a general self-consistency check of the classical background assumption. 
We provide a natural generalization to local QFT in curved spacetime, and we discuss why such self-consistency check is different from perturbative calculations of graviton-scattering.

@@

Double-slit interference is the signature experiment to demonstrate the nature of quantum mechanics. Figure 1 shows a simple example of such experiment.
When a group of particles of the same momentum pass through slit 1 only, they reach the final screen as some probability distribution $P_1(y)$. 
When they pass through slit 2 only, they reach the final screen as a probability distribution $P_2(y)$.
When both slits are open, the probability distribution we find on the final screen is not a simple sum; $P_{12}(y) \neq P_1(y) + P_2(y)$,
instead, the final screen shows an interference pattern which can only be explained by wavefunctions.
$P_i = \langle \phi_i | \phi_i \rangle$ and $P_{12} = (\langle \phi_1|+\langle \phi_2| ) (|\phi_1\rangle + | \phi_2 \rangle)=P_1 + P_2 + 2{\rm Re}\langle\phi_1|\phi_2\rangle$.

@@

This standard explanation works because we assume the particle is the only quantum-mechanical system here, and it can be describe by a pure state, $|\phi\rangle_{\rm particle} = |\phi_1\rangle + |\phi_2\rangle$.
But if quantum mechanics is correct, this particle is not the only thing to be described by a wavefunction. 
For example, we can also describe the double-slit by its wavefunction $|\psi\rangle_{\rm ds}$. The total wavefunction of the combined system can be in a pure state,
\begin{equation}
|\Psi\rangle_{\rm combined} = |\psi_1\rangle_{\rm ds}|\phi_1\rangle_{\rm particle} 
+ |\psi_2\rangle_{\rm ds}|\phi_2\rangle_{\rm particle} ~,
\end{equation}
but either subsystem does not have to be pure.

@@

\begin{figure}[tb]
\begin{center}.
\includegraphics[scale = 0.5]{bh_coherence.pdf}
\caption{Schematic showing the rule we advocate; as classical coherence fails, we loose quantum unitarity.}
\label{fig-QFT}
\end{center}
\end{figure}

@@

QFT theorists might think that a more direct self-consistency check is feasible.
Instead of parametrizing a classical geometric uncertainty, one can treat deviations from the background geometry as a field and quantize it as any other fields.
Then a standard QFT calculation of scattering with this new field (graviton) should capture all interactions with the geometry, and a small scattering amplitude should guarantee that the interaction is suppressed. 
The concern for this method is that IR issues for these scattering calculations cannot be unambiguously regulated as in global Minkowski space.
Since soft gravitons lead to real physical changes in the geometry that can be seen in memory effects \cite{Wei65,HeLys14}, an IR ambiguity signals our ignorance on how these geometric changes enter QFT.
Thus a direct graviton scattering calculation cannot be the full extent of a self-consistency check, and it is reasonable for our method to give extra constraints on the reliability of local QFT.

@@

The usual treatment of how a near-horizon mode becomes an asymptotic Hawking quantum also uses a fixed-background assumption.
The background is the Schwarzschild geometry.
\begin{equation}
ds^2 = -\left(1-\frac{2M}{r}\right)dt^2 + \frac{dr^2}{1-\frac{2M}{r}} + r^2d\Omega_2^2~.
\label{eq-metric}
\end{equation}
When one applies local quantum field theory to describe a field on this fixed background, the result is by-definition unitary, but that is not necessary a physical fact. 
Given how little we understand quantum gravity, a classical self-consistency check as we described in the previous section is a reasonable thing to do.

>>>>

\begin{abstract}
The black hole information paradox presumes that quantum field theory in curved spacetime can provide unitary propagations from a near-horizon mode to an asymptotic Hawking quantum. 
Instead of invoking conjectural quantum-gravity effects to modify such an assumption, we propose a self-consistency check.
We establish an analogy to Feynman's analysis of a double-slit experiment. 
Feynman showed that unitary propagation of the interfering particles, namely ignoring the entanglement with the double-slit, becomes an arbitrarily reliable assumption when the screen upon which the interference pattern is projected is infinitely far away.
We argue for an analogous self-consistency check for quantum field theory in curved spacetime.
We apply it to the propagation of Hawking quanta and test whether ignoring the entanglement with the geometry also becomes arbitrarily reliable in the limit of a large black hole. 
We present curious results to suggest a negative answer, and we discuss how this loss of na\"ive unitarity in QFT might be related to a solution of the paradox based on the soft-hair-memory effect.
\end{abstract}

@@

The black hole information paradox \cite{Haw76a} was sharpened to show a self-inconsistency among the following three widely-believed statements \cite{AMPS}\footnote{Different physicists may believe in some of those more strongly than the others, but very few would have argued against any single statement without the explicit conflict in the information paradox.}:
\begin{itemize}
\item {\bf Unitary evaporation:} A black hole will totally evaporate; the formation and evaporation of a black hole is described by an asymptotic observer as a unitary $S$-matrix, whose size is the exponential of the black hole's Bekenstein entropy.
\item {\bf General relativity:} The collapse-Schwarzschild geometry given by general relativity is a valid description of the spacetime everywhere except for near the singularity.
Therefore, there is no drama while crossing the horizon.
\item {\bf Quantum field theory in curved spacetime:} Away from the singularity, we can apply local quantum field theory (QFT) which describes microscopic steps of evaporation. 
For large black holes, it manifests itself as unitary propagation of every near-horizon mode into every asymptotic Hawking quantum. The back-reaction of such effect is no more than a Schwarzschild metric with a very slowly changing mass.
\end{itemize}
Basically, Statement Three demands that a near horizon mode is related to an asymptotic Hawking quantum by a unitary transformation, and therefore carries the same qubit of information. 
However, Statement One and Two demand this unique qubit to be purified by two distinct objects, violating monogamy \cite{Bou13}. 
This conflict has inspired many different proposals to modify one of the above three statements. 
For example, information loss or remnant \cite{Bek94} modifies Statement One; 
firewall \cite{BraPir09,AMPS}, ER=EPR \cite{MalSus13}, pre-Hawking radiation \cite{KawMat13,Ho16}, or some version of the fuzzball \cite{Mat15} modifies Statement Two;
various proposals of nonlocal effects near the horizon \cite{Gid12,DodSil15,OsuPag16}, causal patch complementarity \cite{HuiYan13,IlgYan13,LowTho14}, or computational complementarity \cite{HH} modify Statement Three.\footnote{Various versions of complementarity basically claim that QFT breaks down if one applies it to compare quantities which are in-principle not measurable (or difficult to measure) by the same observer.}

In this paper, we will try to provide a stronger motivation to modify Statement Three. 
In fact, we will argue that it is already problematic on its own. 
In other words, {\bf the application of local QFT to unitary propagation of individual Hawking quantum is not self-consistent. }

Before presenting our argument, we will first explain its relation to some existing ideas.
The recent paper by Osuga and Page \cite{OsuPag16} provides an abstract framework of how modifying Statement Three resolves the paradox.
Basically, the near horizon mode is purified by its interior partner as demanded by Statement Two.
However, it propagates out in a non-unitary process, losing its original state and gaining the appropriate information to restore the unitary $S$-matrix demanded by Statement One. 
For this abstract model to work in practice, this non-unitary propagation should be understood as an interaction with a hidden system, which actually exchanges the information within the out-going quantum with a qubit of information in the hidden system.
Here hidden simply means something that a na\"ive application of local QFT is ignorant to.
The obvious candidate of such hidden system is the geometry.

One proposal along this line of thoughts was discussed in \cite{NomVar12}. 
The idea was that the black hole acquires a large uncertainty during the evaporation process. 
If we treat the superposition of different classical geometries as a quantum system, it may qualify as the hidden system that is responsible for the non-unitary propagation of Hawking quanta. 
Unfortunately, the early version of this proposal can be defeated by counting entropy.
In order for such non-unitary process to restore the asymptotic $S$-matrix, the hidden system must store the required information, which is comparable to the amount of microscopic information carried by the black hole.
However, the black-hole-no-hair theorem states that two black-hole geometries with identical macroscopic parameters (mass, momentum, angular momentum, charge) are isometric to each other. 
This limits the number of different classical geometries to be parametrized by macroscopic parameters only.
As a result, the Hilbert space of superposed classical geometries has far fewer dimensions than what is necessary to carry the required amount of information.\footnote{We thank Raphael Bousso for private communication.}
Thus, even if a propagating Hawking quantum does interact with this hidden system, it cannot reproduce a unitary $S$-matrix.
Therefore such modification of Statement Three alone is insufficient to resolve the paradox.

The recent development on black-hole-soft-hairs \cite{HawPer16} hinted a way to revive this idea.
Essentially, it was realized that two isometric (regions of) geometries should not always be considered as the same state.
If one maps the geometries with congruences of geodesics, then two initially identical geometries with identical mappings may evolve into two late-time isometric geometries with different mappings. 
Since changes in the mapping geodesics are measurable memory effects, a consistent physical theory must be able to distinguish them.
Thus isometric regions with different maps should be described as different states.
This realization may dramatically increase the number of different states encoded in classical geometries.
A tentative counting of state on a black hole horizon yields exactly the required number such that a modification to Statement Three can indeed resolve the paradox \cite{HawPer16}.

In this paper, we will take another step that is complementary to the above progress. It is exciting to know that modifying Statement Three can, in-principle, resolve the paradox.
However, it remains a great mystery why local QFT should break down in low curvature regions. 
If we just demand that it breaks down in order to resolve the information paradox, then there is no reason why this solution is preferred over modifying Statement One or Two. 
We would like to argue that there is a self-consistency check that we should apply to QFT in curved spacetime, for a reason that is totally independent from the information paradox.
Applying such self-consistency check, we can show that a unitary propagation from a near horizon mode to an asymptotic Hawking quantum is already self-inconsistent. 
Thus, it is only natural to modify Statement Three.

@@

Fortunately, Feynman, in his famous lectures, presented an interesting trick to circumvent this obstacle. 
This trick, to a certain extent, enables us to perform a {\bf self-consistency check} of the classical-background assumption {\bf without} knowing the quantum nature of the background.
In Sec.\ref{sec-DoubleSlit}, we will review Feynman's analysis of the double-slit experiment. 
In his case, the double-slit is the classical background, the particle passing through it is the quantum subsystem, and subsystem unitarity is checked by whether an interference pattern can be observed on the final projection screen. 
He showed that we can ignore the quantum details of the double-slit and summarize that as an uncertainty in its position, which is a classical quantity.
\footnote{Here we mean that an uncertainty in position is a quantity that one can directly plug into the framework/equations of a classical theory, as oppose to states or operators which can only be understood by a quantum theory.} 
When such uncertainty is fixed, a classical calculation can show that the interference pattern is always visible as long as we put the final projection screen to be infinitely far away. 
It is in this sense that a unitary evolution of the subsystem is an arbitrarily good approximation.

The fundamental principle behind Feynman's trick is that ``classical wave coherence'',
\footnote{We remind the readers that ``coherence'' has a classical (likely original) meaning as superpositions of waves of stably similar phases.}
 in his case the observability of the interference pattern, should be taken as the prerequisite of ``quantum unitarity''. 
In Sec.\ref{sec-QFT}, we argue that this principle enables a general self-consistency check of the classical background assumption. 
We provide a natural generalization to local QFT in curved spacetime, and we discuss why such self-consistency check is different from perturbative calculations of graviton-scattering.

@@

Double-slit interference is the signature experiment to demonstrate the nature of quantum mechanics. Figure 1 shows a simple example of such experiment.
When a group of particles of the same momentum pass through slit 1 only, they reach the final screen as some probability distribution $P_1(y)$. 
When they pass through slit 2 only, they reach the final screen as a probability distribution $P_2(y)$.
When both slits are open, the probability distribution we find on the final screen is not a simple sum; $P_{12}(y) \neq P_1(y) + P_2(y)$.
Instead, the final screen shows an interference pattern which can only be explained by wavefunctions.
$P_i = \langle \phi_i | \phi_i \rangle$ and $P_{12} = (\langle \phi_1|+\langle \phi_2| ) (|\phi_1\rangle + | \phi_2 \rangle)=P_1 + P_2 + 2{\rm Re}\langle\phi_1|\phi_2\rangle$.

@@

This standard explanation assumes the particle is the only quantum-mechanical system here, and it can be describe by a pure state, $|\phi\rangle_{\rm particle} = |\phi_1\rangle + |\phi_2\rangle$.
But if quantum mechanics is correct, this particle is not the only thing to be described by a wavefunction. 
For example, we can also describe the double-slit by its wavefunction $|\psi\rangle_{\rm ds}$. The total wavefunction of the combined system can be in a pure state,
\begin{equation}
|\Psi\rangle_{\rm combined} = |\psi_1\rangle_{\rm ds}|\phi_1\rangle_{\rm particle} 
+ |\psi_2\rangle_{\rm ds}|\phi_2\rangle_{\rm particle} ~,
\end{equation}
but either subsystem does not have to be pure.

@@

\begin{figure}[tb]
\begin{center}.
\includegraphics[scale = 0.5]{bh_coherence.pdf}
\caption{Schematic showing the self-consistency check we advocate. 
If there is a large enough uncertainty in the classical geometry, a classical mode function can have a significant uncertainty in its phases.
If this happens, classical waves lose coherence, and the evolution of their quantum states lose unitarity.}
\label{fig-QFT}
\end{center}
\end{figure}

@@

QFT theorists might think that the more direct self-consistency check is feasible.
Instead of parametrizing a classical geometric uncertainty, one can treat deviations from the background geometry as a field and quantize it as any other fields.
Then a standard QFT calculation of scattering with this new field (graviton) should capture all interactions with the geometry, and a small scattering amplitude should guarantee that the interaction is suppressed. 
The concern for this method is that IR issues for these scattering calculations cannot be unambiguously regulated as in global Minkowski space.
Since soft gravitons lead to real physical changes in the geometry that can be seen in memory effects \cite{Wei65,HeLys14}, an IR ambiguity signals our ignorance on how these geometric changes enter QFT.
Thus a direct graviton scattering calculation cannot be the full extent of a self-consistency check, and it is reasonable for our method to give extra constraints on the reliability of local QFT.

@@

The usual treatment of how a near-horizon mode becomes an asymptotic Hawking quantum also uses a fixed-background assumption.
The background is the Schwarzschild geometry.
\begin{equation}
ds^2 = -\left(1-\frac{2M}{r}\right)dt^2 + \frac{dr^2}{1-\frac{2M}{r}} + r^2d\Omega_2^2~.
\label{eq-metric}
\end{equation}
When one applies local quantum field theory to describe a field on this fixed background, the results always appear to be unitary, but that is not necessary a physical fact. 
Given how little we understand quantum gravity, a classical self-consistency check as we described in the previous section is a reasonable thing to do.
